---
title: "hearst_data_test"
author: "Paul Kiefer"
date: "2026-02-04"
output: html_document
---

LOAD LIBRARIES:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Core & Manipulation
library(tidyverse)
library(lubridate)
library(janitor)
library(stringr)
library(readxl)

# Spatial & Census
library(sf)
library(tigris)
library(tidycensus)
library(tidygeocoder)

# Visualization
library(gridExtra)
library(leaflet)
library(htmltools)
library(scales)


# CONFIGURATION
# Set to TRUE to re-run geocoding (takes time). Set to FALSE to skip if data exists.
RUN_GEOCODING <- TRUE
```

LOAD AND CLEAN DATA:

* INVOLVES CUSTOM CLEANING AND FOOTNOTE-PRESERVING FUNCTIONS*

```{r}

# ------------------------------
# Function to clean a single offenses dataset
# ------------------------------
clean_offenses <- function(path, skip_rows = 3, rename_arson = FALSE) {
  df <- read_xlsx(path, skip = skip_rows) %>%
    clean_names() %>%
    mutate(
      # Extract footnotes from state and city
      state_footnote = str_extract(state, "\\d+"),
      city_footnote  = str_extract(city, "\\d+"),
      
      # Remove footnote numbers
      state = str_remove(state, "\\d+"),
      city  = str_remove(city, "\\d+"),
      
      # Capitalize
      state = str_to_upper(state),
      city  = str_to_upper(city)
    ) %>%
    fill(state, .direction = "down")  # Only fill state structurally

  
  return(df)
}
```

```{r}
# ------------------------------
# Function to build footnotes table from cleaned dataset
# ------------------------------
build_footnotes <- function(df) {
  
  # Footnote rows: only rows with footnote number and text ending with a period
  state_footnote_rows <- df %>%
    filter(!is.na(state_footnote) & str_detect(state, "\\.$")) %>%
    transmute(footnote_number = state_footnote, footnote = state)
  
  city_footnote_rows <- df %>%
    filter(!is.na(city_footnote) & str_detect(city, "\\.$")) %>%
    transmute(footnote_number = city_footnote, footnote = city)
  
  footnote_rows <- bind_rows(state_footnote_rows, city_footnote_rows) %>%
    distinct(footnote_number, footnote)
  
  # All states linked to each footnote
  state_locations <- df %>%
    filter(!is.na(state_footnote) & !str_detect(state, "\\.$")) %>%
    distinct(state_footnote, state) %>%
    group_by(state_footnote) %>%
    summarize(states = paste(sort(unique(state)), collapse = ", "), .groups = "drop") %>%
    rename(footnote_number = state_footnote)
  
  # All cities linked to each footnote
  city_locations <- df %>%
    filter(!is.na(city_footnote) & !str_detect(city, "\\.$")) %>%
    mutate(city_state = paste0(city, ", ", state)) %>%
    distinct(city_footnote, city_state) %>%
    group_by(city_footnote) %>%
    summarize(cities = paste(sort(unique(city_state)), collapse = "; "), .groups = "drop") %>%
    rename(footnote_number = city_footnote)
  
  # Combine footnote text, states, cities
  footnotes <- footnote_rows %>%
    left_join(state_locations, by = "footnote_number") %>%
    left_join(city_locations, by = "footnote_number") %>%
    arrange(as.integer(footnote_number))
  
  return(footnotes)
}


```

```{r}
# ------------------------------
# Apply cleaning and footnote functions to all years
# ------------------------------
offenses_2024 <- clean_offenses(
  "data/fbi-cde-city-offenses/CIUS_Table_8_Offenses_Known_to_Law_Enforcement_by_State_by_City_2024.xlsx",
  skip_rows = 3
)
footnotes_2024 <- build_footnotes(offenses_2024)

offenses_2023 <- clean_offenses(
  "data/fbi-cde-city-offenses/CIUS_Table_8_Offenses_Known_to_Law_Enforcement_by_State_by_City_2023.xlsx",
  skip_rows = 3
)
footnotes_2023 <- build_footnotes(offenses_2023)

offenses_2022 <- clean_offenses(
  "data/fbi-cde-city-offenses/CIUS_Table_8_Offenses_Known_to_Law_Enforcement_by_State_by_City_2022.xlsx",
  skip_rows = 3
)
footnotes_2022 <- build_footnotes(offenses_2022)

offenses_2021 <- clean_offenses(
  "data/fbi-cde-city-offenses/CIUS_Table_8_Offenses_Known_to_Law_Enforcement_by_State_by_City_2021.xlsx",
  skip_rows = 3,
  rename_arson = TRUE
)
footnotes_2021 <- build_footnotes(offenses_2021)

offenses_2020 <- clean_offenses(
  "data/fbi-cde-city-offenses/CIUS_Table_8_Offenses_Known_to_Law_Enforcement_by_State_by_City_2020.xlsx",
  skip_rows = 5
)
footnotes_2020 <- build_footnotes(offenses_2020)
```

JOIN OFFENSE TABLES 

* STANDARDIZE COLUMNS AND PRESERVE YEARS *

```{r}
# ------------------------------
# Helper function to standardize column names
# ------------------------------
standardize_offense_columns <- function(df) {
  df %>%
    # Remove any trailing numbers or underscores from column names
    rename_with(~ str_remove_all(., "_?\\d+$")) %>%
    
    # Explicitly rename total columns if they exist
    rename_with(~ ifelse(. == "violent_crime", "violent_crime_total", .)) %>%
    rename_with(~ ifelse(. == "property_crime", "property_crime_total", .))
}


# ------------------------------
# Helper function to prepare a single offenses table
# ------------------------------
prepare_offenses <- function(df, year) {
  df %>%
    select(-state_footnote, -city_footnote) %>%  # remove footnote columns
    mutate(year = year) %>%                      # add year column
    standardize_offense_columns()                # standardize column names
}

# ------------------------------
# Apply to each year
# ------------------------------
offenses_2024_clean <- prepare_offenses(offenses_2024, 2024)
offenses_2023_clean <- prepare_offenses(offenses_2023, 2023)
offenses_2022_clean <- prepare_offenses(offenses_2022, 2022)
offenses_2021_clean <- prepare_offenses(offenses_2021, 2021)
offenses_2020_clean <- prepare_offenses(offenses_2020, 2020)

# ------------------------------
# Combine into a single dataset
# ------------------------------
offenses_all_years <- bind_rows(
  offenses_2020_clean,
  offenses_2021_clean,
  offenses_2022_clean,
  offenses_2023_clean,
  offenses_2024_clean
)
```

PULL ACS DATA FOR 2020-2024 AT CITY LEVEL 

* FOCUS ON POPULATION, AGE, INCOME AND HOUSING STATS *

```{r}

#--------------------------------------------------------
# FETCH & CLEAN CENSUS (ACS) DATA
#--------------------------------------------------------
# Years to pull
acs_years <- 2020:2024

# Variables to pull (these hold across 2020–2024)
acs_vars <- list(
  population    = "B01001_001",
  median_age    = "B01002_001",
  median_income = "B06011_001",
  renter_units  = "B25012_010",
  owner_units   = "B25012_002",
  vacant_units  = "B25002_003"
)

# -----------------------------
# Function: Safe Variable Fetch
# -----------------------------
# Pulls a single variable (e.g., median income) for a specific year.
# Returns a clean dataframe with just GEOID, NAME, and the variable.
get_acs_variable <- function(acs_year, var_name, var_code) {
  message("Pulling ", var_name, " for ACS year: ", acs_year)
  
  df <- get_acs(
    geography = "place",
    variables = c(temp = var_code),
    year = acs_year,
    survey = "acs5",
    geometry = FALSE,
    output = "wide"
  ) %>%
    rename(!!var_name := tempE) %>%
    select(GEOID, NAME, !!var_name)
  
  return(df)
}

# -----------------------------
# Function: Fetch All Vars for One Year
# -----------------------------
# Loops through all variables in `acs_vars`, fetches them, and joins them.
# Also cleans the city/state names to match FBI format (UPPERCASE).
get_city_acs_year <- function(acs_year) {
  # 1. Fetch each variable individually
  dfs <- lapply(names(acs_vars), function(var_name) {
    get_acs_variable(acs_year, var_name, acs_vars[[var_name]])
  })
  
  # 2. Merge all variable dataframes by GEOID
  df_year <- Reduce(function(x, y) full_join(x, y, by = c("GEOID", "NAME")), dfs)
  
  # 3. Clean Geography Names
  # Splits "Portland city, Oregon" -> City: "PORTLAND", State: "OREGON"
  df_year <- df_year %>%
    separate(NAME, into = c("city_raw", "state_name"), sep = ", ", extra = "merge") %>%
    mutate(
      city = str_remove(city_raw, "\\s+(town|city|village|CDP)$") %>% str_to_upper(),
      state_name = str_to_upper(state_name)
    ) %>%
    select(-city_raw) %>%
    # 4. Attach State FIPS codes (critical for accurate joining later)
    left_join(
      tigris::states(cb = TRUE) %>%
        st_drop_geometry() %>%
        select(state_name = NAME, state_fips = STATEFP) %>%
        mutate(state_name = str_to_upper(state_name)),
      by = "state_name"
    )
  
  df_year <- df_year %>% mutate(acs_year = acs_year)
  
  return(df_year)
}

# -----------------------------
# Execute Fetch Loop
# -----------------------------
# This runs the functions above for every year in your range
acs_data_list <- lapply(acs_years, get_city_acs_year)
names(acs_data_list) <- acs_years


#----------------------------------------------------------
# SECTION 2: CALCULATE DERIVED METRICS & JOIN TO CRIME DATA
# ---------------------------------------------------------

# -----------------------------
# Step 1: Prepare ACS data
# -----------------------------
# Calculates housing percentages (Owner/Renter/Vacant) before merging
acs_data_prepped <- lapply(acs_data_list, function(df) {
  df %>%
    mutate(
      total_housing_units = renter_units + owner_units + vacant_units,
      pct_renter = if_else(total_housing_units > 0, renter_units / total_housing_units * 100, NA_real_),
      pct_owner  = if_else(total_housing_units > 0, owner_units  / total_housing_units * 100, NA_real_),
      pct_vacant = if_else(total_housing_units > 0, vacant_units / total_housing_units * 100, NA_real_)
    ) %>%
    # Explicitly select state_name to ensure it survives for the join
    select(city, state_name, state_fips, acs_year,
           population, median_age, median_income,
           total_housing_units, pct_renter, pct_owner, pct_vacant)
}) 

# Combine all years into one master ACS dataframe
acs_all_years <- bind_rows(acs_data_prepped)

# -----------------------------
# Step 2: Join ACS to Offenses
# -----------------------------
# Merges Census data with FBI Crime data based on City, State, and Year

# Safety check: Ensure population is numeric
acs_all_years <- acs_all_years %>%
  mutate(population = as.numeric(population))

offenses_with_acs <- offenses_all_years %>%
  # Join operation
  left_join(
    acs_all_years %>% rename(acs_population = population), # Rename to avoid collisions
    by = c(
      "city",
      "state" = "state_name", 
      "year"  = "acs_year"
    )
  ) %>%
  # Calculate per-capita crime rates (per 1,000 residents)
  mutate(
    violent_crime_total  = as.numeric(violent_crime_total),
    property_crime_total = as.numeric(property_crime_total),
    
    violent_crime_per_1000  = violent_crime_total / acs_population * 1000,
    property_crime_per_1000 = property_crime_total / acs_population * 1000,
    arson_per_1000          = arson / acs_population * 1000,
    robbery_per_1000        = robbery / acs_population * 1000,
    rape_per_1000           = rape / acs_population * 1000
  )
```

JOIN WITH OFFENSE DATA TO CREATE PER CAPITA FIGURES 

* ALSO JOIN AGE, INCOME AND HOUSING DATA *

```{r}
# -----------------------------
# Step 1: Prepare ACS data
# -----------------------------
acs_data_prepped <- lapply(acs_data_list, function(df) {
  df %>%
    mutate(
      total_housing_units = renter_units + owner_units + vacant_units,
      pct_renter = if_else(total_housing_units > 0, renter_units / total_housing_units * 100, NA_real_),
      pct_owner  = if_else(total_housing_units > 0, owner_units  / total_housing_units * 100, NA_real_),
      pct_vacant = if_else(total_housing_units > 0, vacant_units / total_housing_units * 100, NA_real_)
    ) %>%
    # Add state_name here so it survives the selection
    select(city, state_name, state_fips, acs_year,
           population, median_age, median_income,
           total_housing_units, pct_renter, pct_owner, pct_vacant)
}) 

# Combine into one dataframe
acs_all_years <- bind_rows(acs_data_prepped)

# -----------------------------
# Step 2: Join ACS to offenses (FIXED)
# -----------------------------

# First, let's verify columns aren't character types
acs_all_years <- acs_all_years %>%
  mutate(population = as.numeric(population))

offenses_with_acs <- offenses_all_years %>%
  # Rename ACS population to avoid collision with FBI population
  left_join(
    acs_all_years %>% rename(acs_population = population), 
    by = c(
      "city",
      "state" = "state_name", 
      "year"  = "acs_year"
    )
  ) %>%
  mutate(
    # Ensure numeric types for the crime columns just in case
    violent_crime_total = as.numeric(violent_crime_total),
    property_crime_total = as.numeric(property_crime_total),
    
    # Use the specific ACS population column we renamed above
    violent_crime_per_1000  = violent_crime_total / acs_population * 1000,
    property_crime_per_1000 = property_crime_total / acs_population * 1000,
    arson_per_1000          = arson / acs_population * 1000,
    robbery_per_1000        = robbery / acs_population * 1000,
    rape_per_1000           = rape / acs_population * 1000
  )

```


CALCULATE CHANGES BETWEEN 2020-2024 AND IDENTIFY TOP 10 CHANGES (PROPERTY AND VIOLENT) IN EACH POPULATION CATEGORY

* POPULATION CATEGORIES ARE ARBITRARY *

```{r}
# -----------------------------
# Step 1: Prepare the 2020 vs 2024 comparison (FIXED)
# -----------------------------
crime_change <- offenses_with_acs %>%
  filter(year %in% c(2020, 2024)) %>%
  
  # --- NEW FIX: Handle Duplicates ---
  # If a city appears twice (e.g. Town vs City), keep the one with the larger population
  group_by(city, state, year) %>%
  slice_max(acs_population, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  # ----------------------------------

  select(city, state, year, acs_population, violent_crime_per_1000, property_crime_per_1000) %>%
  
  # Pivot wider to get 2020 and 2024 on the same row
  pivot_wider(
    id_cols = c(city, state),
    names_from = year,
    values_from = c(acs_population, violent_crime_per_1000, property_crime_per_1000),
    names_sep = "_"
  ) %>%
  
  # Remove cities that don't have data for both years
  drop_na(violent_crime_per_1000_2020, violent_crime_per_1000_2024) %>%
  
  mutate(
    # Calculate absolute change in rates
    violent_change  = violent_crime_per_1000_2024 - violent_crime_per_1000_2020,
    property_change = property_crime_per_1000_2024 - property_crime_per_1000_2020,
    
    # Create the 4 categories based on 2024 population
    # Note: We use acs_population_2024 (the column created by pivot_wider)
    pop_category = case_when(
      acs_population_2024 < 50000  ~ "1. Under 50k",
      acs_population_2024 < 200000 ~ "2. 50k to 200k",
      acs_population_2024 < 700000 ~ "3. 200k to 700k",
      TRUE                         ~ "4. Over 700k"
    )
  )

# -----------------------------
# Step 2: Function to extract top/bottom 10 (FIXED)
# -----------------------------
get_top_changes <- function(data, change_col, base_rate_col, direction = "increase") {
  
  # Construct the 2020 and 2024 column names manually
  col_2020 <- paste0(base_rate_col, "_2020")
  col_2024 <- paste0(base_rate_col, "_2024")
  
  data %>%
    filter(!is.na(pop_category)) %>%
    group_by(pop_category) %>%
    # Sort by the change column
    slice_max(order_by = if(direction == "increase") !!sym(change_col) else -!!sym(change_col), n = 10) %>%
    mutate(rank = row_number()) %>%
    ungroup() %>%
    # Select columns using the explicit names we constructed
    select(
      pop_category, 
      rank, 
      city, 
      state, 
      start_rate = !!sym(col_2020),
      end_rate   = !!sym(col_2024),
      change     = !!sym(change_col)
    ) %>%
    arrange(pop_category, rank)
}

# -----------------------------
# Step 3: Generate the 4 Lists
# -----------------------------

# 1. Violent Crime Increases
violent_increases <- get_top_changes(crime_change, "violent_change", "violent_crime_per_1000", "increase")

# 2. Violent Crime Decreases
violent_decreases <- get_top_changes(crime_change, "violent_change", "violent_crime_per_1000", "decrease")

# 3. Property Crime Increases
property_increases <- get_top_changes(crime_change, "property_change", "property_crime_per_1000", "increase")

# 4. Property Crime Decreases
property_decreases <- get_top_changes(crime_change, "property_change", "property_crime_per_1000", "decrease")

```

MAP CITIES IN TOP 10 FOR EACH CATEGORY

* CATEGORIES ARE VIOLENT/INCREASE, VIOLENT/DECREASE, PROPERTY/INCREASE, AND PROPERTY/DECREASE *

```{r}
# ---------------------------------------------------------
# Function 1: Geocode the Cities (Find Lat/Long)
# ---------------------------------------------------------
add_coordinates <- function(df) {
  # check if df is empty
  if(nrow(df) == 0) return(df)
  
  message("Geocoding ", nrow(df), " cities... please wait.")
  
  df %>%
    # Create a search string for the geocoder
    mutate(search_query = paste(city, state, "USA", sep = ", ")) %>%
    # Use OSM (Nominatim) service - it is free and requires no key
    geocode(address = search_query, method = 'osm', lat = lat, long = long) %>%
    drop_na(lat, long) # Drop any that failed to locate
}

# ---------------------------------------------------------
# Function 2: Create the Interactive Map
# ---------------------------------------------------------
create_crime_map <- function(data, title_text, metric_label) {
  
  if(nrow(data) == 0) {
    message("No data available for: ", title_text)
    return(NULL)
  }
  
  # Define a color palette for the 4 Population Categories
  # (Yellow -> Orange -> Red -> Purple)
  pal <- colorFactor(
    palette = c("#f1c40f", "#e67e22", "#e74c3c", "#8e44ad"),
    domain = data$pop_category
  )
  
  leaflet(data) %>%
    addProviderTiles(providers$CartoDB.Positron) %>% # Clean grey background
    
    # Add the Circle Markers
    addCircleMarkers(
      lng = ~long, 
      lat = ~lat,
      radius = 8,
      color = ~pal(pop_category),
      stroke = TRUE, fillOpacity = 0.7, weight = 1,
      
      # The Tooltip (Hover)
      label = ~paste0(rank, ". ", city, ", ", state),
      
      # The Popup (Click)
      popup = ~paste0(
        "<strong>", city, ", ", state, "</strong><br/>",
        "Category: ", pop_category, "<br/>",
        "Rank: ", rank, "<br/><hr>",
        "2020 Rate: ", round(start_rate, 1), "<br/>",
        "2024 Rate: ", round(end_rate, 1), "<br/>",
        "<strong>Change: ", round(change, 1), "</strong>"
      )
    ) %>%
    
    # Add a Legend
    addLegend(
      "bottomright",
      pal = pal,
      values = ~pop_category,
      title = "Population Category",
      opacity = 1
    ) %>%
    
    # Add a Title Control (Custom HTML)
    addControl(
      html = paste0("<h3 style='margin:0; padding:10px; background:white; opacity:0.9;'>", 
                    title_text, "</h3>"),
      position = "topright"
    )
}

# ---------------------------------------------------------
# EXECUTION: Geocode the 4 Lists
# ---------------------------------------------------------
# (This assumes you have already run the previous step to create these lists)
message("Starting geocoding process...")

map_data_violent_inc <- add_coordinates(violent_increases)
map_data_violent_dec <- add_coordinates(violent_decreases)
map_data_prop_inc    <- add_coordinates(property_increases)
map_data_prop_dec    <- add_coordinates(property_decreases)

message("Geocoding complete.")

# ---------------------------------------------------------
# VIEW MAPS
# ---------------------------------------------------------

# Map 1: Violent Crime Increases
create_crime_map(
  map_data_violent_inc, 
  title_text = "Top Increases: Violent Crime (2020-2024)",
  metric_label = "Violent Crimes per 1,000"
)

# Map 2: Violent Crime Decreases
create_crime_map(
  map_data_violent_dec, 
  title_text = "Top Decreases: Violent Crime (2020-2024)",
  metric_label = "Violent Crimes per 1,000"
)

# Map 3: Property Crime Increases
create_crime_map(
  map_data_prop_inc, 
  title_text = "Top Increases: Property Crime (2020-2024)",
  metric_label = "Property Crimes per 1,000"
)

# Map 4: Property Crime Decreases
create_crime_map(
  map_data_prop_dec, 
  title_text = "Top Decreases: Property Crime (2020-2024)",
  metric_label = "Property Crimes per 1,000"
)
```
CALCULATE CORRELATION BETWEEN DEMOGRAPHIC CHANGE (POP, MEDIAN INCOME, AGE, HOUSING UNITS, RENTER SHARE) AND CHANGE IN VIOLENT/PROPERTY CRIME

```{r}
# -----------------------------
# 1. Prepare Analysis Data
# -----------------------------
# Calculate change for all ACS vars and both crime types
analysis_data <- offenses_with_acs %>%
  filter(year %in% c(2020, 2024)) %>%
  mutate(housing_units_per_capita = total_housing_units / acs_population) %>%
  
  # Deduplicate: Keep the row with the largest population per city/year
  group_by(city, state, year) %>%
  slice_max(acs_population, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  
  select(city, state, year, acs_population, median_income, median_age, pct_renter, 
         housing_units_per_capita, violent_crime_per_1000, property_crime_per_1000) %>%
  
  pivot_wider(
    id_cols = c(city, state),
    names_from = year,
    values_from = c(acs_population, median_income, median_age, pct_renter, 
                    housing_units_per_capita, violent_crime_per_1000, property_crime_per_1000),
    names_sep = "_"
  ) %>%
  drop_na() %>%
  mutate(
    # Independent Variables (Predictors)
    pop_change_pct            = (acs_population_2024 - acs_population_2020) / acs_population_2020,
    income_change_pct         = (median_income_2024 - median_income_2020) / median_income_2020,
    age_change                = median_age_2024 - median_age_2020,
    housing_per_capita_change = housing_units_per_capita_2024 - housing_units_per_capita_2020,
    pct_renter_change         = pct_renter_2024 - pct_renter_2020,
    
    # Dependent Variables (Outcomes)
    violent_change_rate       = violent_crime_per_1000_2024 - violent_crime_per_1000_2020,
    property_change_rate      = property_crime_per_1000_2024 - property_crime_per_1000_2020
  )

# Define variables for labels
vars_to_test <- list(
  "pop_change_pct"            = "Population Change (%)",
  "income_change_pct"         = "Median Income Change (%)",
  "age_change"                = "Change in Median Age",
  "housing_per_capita_change" = "Change in Housing Units per Capita",
  "pct_renter_change"         = "Change in % Renter"
)

# -----------------------------
# 2. Violent Crime Analysis
# -----------------------------
message("--- Violent Crime Correlations ---")
cor_matrix_violent <- analysis_data %>%
  select(pop_change_pct, income_change_pct, age_change, 
         housing_per_capita_change, pct_renter_change, violent_change_rate) %>%
  cor(use = "complete.obs")

print(cor_matrix_violent["violent_change_rate", ])

# Violent Crime Plots
plots_violent <- lapply(names(vars_to_test), function(var) {
  ggplot(analysis_data, aes(x = .data[[var]], y = violent_change_rate)) +
    geom_point(alpha = 0.5, color = "#2c3e50") +
    geom_smooth(method = "lm", color = "#e74c3c", se = FALSE) + # Red line for violent
    labs(title = vars_to_test[[var]], y = "Change in Violent Crime Rate", x = NULL) +
    theme_minimal()
})

# Display Violent Grid
do.call(grid.arrange, c(plots_violent, ncol = 2, top = "Correlates of Violent Crime Change (2020-2024)"))


# -----------------------------
# 3. Property Crime Analysis
# -----------------------------
message("--- Property Crime Correlations ---")
cor_matrix_prop <- analysis_data %>%
  select(pop_change_pct, income_change_pct, age_change, 
         housing_per_capita_change, pct_renter_change, property_change_rate) %>%
  cor(use = "complete.obs")

print(cor_matrix_prop["property_change_rate", ])

# Property Crime Plots
plots_prop <- lapply(names(vars_to_test), function(var) {
  ggplot(analysis_data, aes(x = .data[[var]], y = property_change_rate)) +
    geom_point(alpha = 0.5, color = "#2c3e50") +
    geom_smooth(method = "lm", color = "#27ae60", se = FALSE) + # Green line for property
    labs(title = vars_to_test[[var]], y = "Change in Property Crime Rate", x = NULL) +
    theme_minimal()
})

# Display Property Grid
do.call(grid.arrange, c(plots_prop, ncol = 2, top = "Correlates of Property Crime Change (2020-2024)"))
```


IDENTIFY OUTLIERS (FAST GROWTH/RISING VIOLENT CRIME AND POPULATION DECLINE/DECLINING CRIME)
```{r}
# -----------------------------
# 1. Define Thresholds (Top/Bottom 25%)
# -----------------------------
# We use quantiles to define "High Growth" and "Big Drop" relative to the dataset
pop_thresholds   <- quantile(analysis_data$pop_change_pct, probs = c(0.25, 0.75))
crime_thresholds <- quantile(analysis_data$violent_change_rate, probs = c(0.25, 0.75))

# High Growth = Top 25% of population change
# High Crime Spike = Top 25% of violent rate increase
# Shrinking = Bottom 25% of population change
# Big Crime Drop = Bottom 25% of violent rate change (largest decreases)

# -----------------------------
# 2. Extract "Boom Spikes" (The Growth Paradox)
# -----------------------------
boom_spikes <- analysis_data %>%
  filter(
    pop_change_pct > pop_thresholds[2],       # High Growth
    violent_change_rate > crime_thresholds[2] # High Crime Spike
  ) %>%
  arrange(desc(violent_change_rate)) %>%      # Sort by worst crime spike
  select(city, state, 
         pop_2024 = acs_population_2024,
         pop_change_pct, 
         housing_change = housing_per_capita_change,
         violent_rate_2020 = violent_crime_per_1000_2020,
         violent_rate_2024 = violent_crime_per_1000_2024,
         violent_change_rate) %>%
  mutate(
    # Format for readability
    pop_change_pct = scales::percent(pop_change_pct, accuracy = 0.1),
    violent_change_rate = round(violent_change_rate, 2)
  )

# -----------------------------
# 3. Extract "Turnaround Towns" (The Shrinkage Paradox)
# -----------------------------
turnaround_towns <- analysis_data %>%
  filter(
    pop_change_pct < pop_thresholds[1],       # Shrinking
    violent_change_rate < crime_thresholds[1] # Big Crime Drop
  ) %>%
  arrange(violent_change_rate) %>%            # Sort by biggest drop (most negative)
  select(city, state, 
         pop_2024 = acs_population_2024,
         pop_change_pct, 
         housing_change = housing_per_capita_change,
         violent_rate_2020 = violent_crime_per_1000_2020,
         violent_rate_2024 = violent_crime_per_1000_2024,
         violent_change_rate) %>%
  mutate(
    pop_change_pct = scales::percent(pop_change_pct, accuracy = 0.1),
    violent_change_rate = round(violent_change_rate, 2)
  )

# -----------------------------
# 4. View Results
# -----------------------------
message("--- BOOM SPIKES: Fast Growth + Rising Crime ---")
print(head(boom_spikes, 10))

message("\n--- TURNAROUND TOWNS: Shrinking Pop + Falling Crime ---")
print(head(turnaround_towns, 10))
```

MAP OUTLIERS BY CATEGORY

```{r}
# Use the same add_coordinates function from before
# (Ensure tidygeocoder is loaded)
library(tidygeocoder)

message("Geocoding Boom Spikes...")
map_data_boom <- add_coordinates(boom_spikes)

message("Geocoding Turnaround Towns...")
map_data_turnaround <- add_coordinates(turnaround_towns)

create_paradox_map <- function(data, title_text, paradox_type = "boom") {
  
  if(nrow(data) == 0) return(NULL)

  # Color Palette
  pal <- colorFactor(
    palette = c("#f1c40f", "#e67e22", "#e74c3c", "#8e44ad"),
    domain = data$pop_category
  )

  leaflet(data) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addCircleMarkers(
      lng = ~long, lat = ~lat, radius = 7,
      color = ~pal(pop_category),
      stroke = TRUE, fillOpacity = 0.8, weight = 1,
      
      label = ~paste0(city, ", ", state),
      
      popup = ~paste0(
        "<strong>", city, ", ", state, "</strong><br/>",
        "Rank: ", rank, "<br/>",
        "Category: ", pop_category, "<br/><hr>",
        # Show Pop Change formatted as %
        "Pop Change: <strong>", scales::percent(pop_change_pct, accuracy = 0.1), "</strong><br/>",
        "Crime Rate (2020): ", round(violent_rate_2020, 1), "<br/>",
        "Crime Rate (2024): ", round(violent_rate_2024, 1), "<br/>",
        "<strong>Crime Change: ", round(violent_change_rate, 2), "</strong>"
      )
    ) %>%
    addLegend("bottomright", pal = pal, values = ~pop_category, title = "Population Size") %>%
    addControl(
      html = paste0("<h3 style='margin:0; padding:10px; background:white; opacity:0.9;'>", title_text, "</h3>"),
      position = "topright"
    )
}

# -----------------------------
# Generate the Maps
# -----------------------------

# Map 1: Boom Spikes (Growing Fast, Crime Spiking)
create_paradox_map(map_data_boom, "Boom Spikes: Fast Growth + Crime Spike")

# Map 2: Turnaround Towns (Shrinking, Crime Falling)
create_paradox_map(map_data_turnaround, "Turnaround Towns: Shrinking + Safer")
```


